A hazards approach to the biometric analysis of infant mortality

### 2023-03-25 paper2.Rmd
add exogenous hazard on normal time scale

We suggest a variation of the Bourgeois-Pichat biometric analysis of infant mortality. Instead of assuming that the cumulative mortality in the last eleven months of infancy follows a uniform distribution, given a log-cube transformation of age, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. This assumption is more satisfactory from a theoretical point of view, since it models the conditional probability of dying. The growth in birth weight also follows this distribution [@bp51a; @bp51b]. In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies. The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs. We give examples of the two methods using data for 19th and 10th century norhern Sweden, where both mehods fit very well. In addition, we give examples from eighteenth Sweden, where the curve is concave during the last eleven month of infancy, despite high levels of mortality in the first month of life.

@@

```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.height = 7}
x <- seq(0, 365, length = 366)
plot(x, g(x), type = "l", xlab = "Days", ylab = "g(Days)", 
     col = "blue", axes = FALSE)
att <- c(0, 28, 183, 365)
axis(1, at = att, labels = att)
axis(2, las = 1, at = g(att), labels = c("0", "g(28)", "g(183)", "365"))
box()
lines(c(0, 365), c(0, 365), lty = 2)
abline(v = c(0, 365), h = c(0, 365))
abline(v = 28, h = g(28), col = "darkgreen", lty = 3)
```

@@

@@

However, this data set is known to be of very high quality [@tbcl99]. Sex ratio at birth 
is as expected and the teonatal mortality is high for all social groups. We are therefore inclined to 
think that the biometric model as it is described here simply is inappropriate during the eighteenth century.
Specifically, the log-cube transformation, which fits well with 19th century data, seems
to be completely out of order in the eighteenth century.

@@

```{r skum, fig.cap = "Skellefteå 1821--1838."}
library(skum)
births <- obs[obs$starttyp == 2 &
                 obs$foddat > as.Date("1820-12-31") &
                 obs$foddat <= as.Date("1838-12-31"), ]
births$event <- births$sluttyp == 2
births$enter <- 365 * births$enter
births$exit <- 365 * births$exit
addup <- births$exit <= 0
births$exit[addup] <- 0.25
infa <- age.window(births, c(0, 365))
##infa$enter <- g(infa$enter)
##infa$exit <- g(infa$exit)
pf(infa)
```

@@

If anything, the hazards method never performs worse than the B-P method.
But the real strength of the hazards method is that it fits naturally into general 
modern survival analysis with censored and truncated data, and also proportional 
hazards models with covariates.

On the other hand, the whole business with the 
biometric analysis and the log-cube transform seems to be highly questionable. 
It works for data in approximately the same time period and similar environment, 
but beyond that, it is not generalizable at all. These methods should be used with 
great care, if at all.

>>>>

We suggest a variation of the Bourgeois-Pichat biometric analysis of infant mortality. Instead of assuming that the cumulative mortality in the last eleven months of infancy follows a uniform distribution, given a log-cube transformation of age, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. This assumption is more satisfactory from a theoretical point of view, since it models the conditional probability of dying. The growth in birth weight also follows this distribution [@bp51a; @bp51b]. In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies. The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs. We give examples of the two methods using data for 19th and 20th century norhern Sweden, where both mehods fit very well. In addition, we give examples from 18th century Sweden, where the curve is concave during the last eleven month of infancy, despite high levels of mortality in the first month of life.

@@

```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.height = 7}
x <- seq(0, 365, length = 366)
plot(x[1:4], g(x[1:4]), type = "l", xlab = "Days", ylab = "g(Days)", 
     col = "blue", axes = FALSE)
att <- c(0, 28, 183, 365)
axis(1, at = att, labels = att)
axis(2, las = 1, at = g(att), labels = c("0", "g(28)", "g(183)", "365"))
box()
lines(c(0, 365), c(0, 365), lty = 2)
abline(v = c(0, 365), h = c(0, 365))
abline(v = 28, h = g(28), col = "darkgreen", lty = 3)
```

@@

And so on.

## Biometric analysis on the natural time scale

Given that the exogeneous mortality follows an Exponential distribution on the log-cube scale, it follows that the cumulative hazards function on the natural time scale is

\begin{equation*}
H(t; \lambda) = \lambda \log^3(t + 1), \; t > 0,
\end{equation*}

and it follows that the hazard function $h$ is given by

\begin{equation*}
h(t; \lambda) = \frac{3 \log^2(t + 1)}{t + 1}, \; t > 0,
\end{equation*}

see Figure \@ref(fig:normal).

```{r normal, fig.cap = "Exogeneos hazard function on the normal time scale."}
x <- 0:365
y <- dg(x)
plot(x, y, type = "l", xlab = "Day", ylab = "Hazards",
     axes = FALSE, col = "blue", lwd = 2)
axis(1, at = c(0, 28, 183, 365))
axis(2, las = 1, at = 0)
box()
abline(h = 0, v = 0)
abline(v = 28, lty = 2, col = "darkgreen")
```

We. may be relatively confident that the shape is OK in the post-neonatal period, where data are available, but the shape during the neonatal period is a pure guess.
Examining data with *known* causes of death might be enlightening.

@@

However, this data set is known to be of very high quality [@tbcl99]. Sex ratio at birth 
is as expected and the neonatal mortality is high for all social groups. We are therefore inclined to 
think that the biometric model as it is described here simply is inappropriate during the eighteenth century.
Specifically, the log-cube transformation, which fits well with 19th century data, seems
to be completely out of order in the eighteenth century.

@@

```{r skum, fig.cap = "Skellefteå 1821--1838."}
library(skum)
births <- obs[obs$starttyp == 2 &
                 obs$foddat < as.Date("1839-01-01") &
                 obs$foddat > as.Date("1820-12-31"), ]
births$event <- births$sluttyp == 2
births$enter <- 365 * births$enter
births$exit <- 365 * births$exit
addup <- births$exit <= 0
births$exit[addup] <- 0.25
infa <- age.window(births, c(0, 365))
##infa$enter <- g(infa$enter)
##infa$exit <- g(infa$exit)
pf(infa)
```

@@

If anything, the hazards method never performs worse than the B-P method.
But the real strength of the hazards method is that it fits naturally into general 
modern survival analysis with censored and truncated data, and also proportional 
hazards models with covariates. 
<!--
On the other hand, the whole business with the 
biometric analysis and the log-cube transform seems to be highly questionable. 
It works for data in approximately the same time period and similar environment, 
but beyond that, it is not generalizable at all. These methods should be used with 
great care, if at all.
-->
Apparently, it works for many places in the mid-twentieth century, but beyond that, it is not generalizable. In fact, also when post-neonatal mortality is following a straight line on the log-cube scale, mortality may well deviate during different parts of the year. This means that the methods should be used with great care and only as a starting point.
